---
title: "Albedo: Surface Albedo Simulation and Analysis"
author: "Karen Dworsky"
date: "2025-11-06"
format: html
execute:
  cache: true
---

## Setup

```{r}
library(dplyr)
library(ggplot2)
library(lubridate)

ASOS <- read.csv("data/ASOS2.txt", header = TRUE, stringsAsFactors = FALSE)

ASOS <- ASOS %>%
  mutate(
    datetime   = as.POSIXct(valid, format = "%Y-%m-%d %H:%M", tz = "UTC"),
    p01i       = as.numeric(ifelse(p01i == "M" | p01i == "" | is.na(p01i), NA, p01i)),
    tmpf       = as.numeric(ifelse(tmpf == "M" | tmpf == "" | is.na(tmpf), NA, tmpf)),
    sknt       = as.numeric(ifelse(sknt == "M" | sknt == "" | is.na(sknt), NA, sknt)),
    temp_c     = (tmpf - 32) * 5/9,
    p01m       = p01i * 25.4,
    wind_ms    = sknt * 0.514444
  ) %>%
  arrange(datetime)
```

## 1. Albedo Simulation

Snow albedo decays exponentially between snowfall events. Fresh snowfall resets albedo to the
maximum value; the decay rate is faster during melt conditions (T > 0°C).

```{r}
alpha_fresh        <- 0.84
alpha_decay_no_melt <- 0.70
alpha_decay_melt   <- 0.50
alpha_soil         <- 0.20
k_no_melt          <- 0.01
k_melt             <- 0.05

asos_data <- ASOS %>%
  mutate(
    snowfall = !is.na(p01i) & p01i > 0 & !is.na(tmpf) & tmpf <= 35,
    albedo   = NA_real_,
    snow_present = FALSE
  )

asos_data$albedo[1]      <- alpha_soil
asos_data$snow_present[1] <- FALSE

for (i in 2:nrow(asos_data)) {
  if (isTRUE(asos_data$snowfall[i])) {
    asos_data$albedo[i]      <- alpha_fresh
    asos_data$snow_present[i] <- TRUE
  } else if (isTRUE(asos_data$snow_present[i - 1])) {
    asos_data$snow_present[i] <- TRUE
    time_diff <- as.numeric(difftime(asos_data$datetime[i],
                                     asos_data$datetime[i - 1], units = "hours"))
    is_melting <- !is.na(asos_data$tmpf[i]) && asos_data$tmpf[i] > 32
    if (is_melting) {
      k         <- k_melt
      alpha_min <- alpha_decay_melt
    } else {
      k         <- k_no_melt
      alpha_min <- alpha_decay_no_melt
    }
    alpha_prev <- asos_data$albedo[i - 1]
    asos_data$albedo[i] <- alpha_min + (alpha_prev - alpha_min) * exp(-k * time_diff)

    if (!is.na(asos_data$tmpf[i]) && asos_data$tmpf[i] > 40) {
      asos_data$snow_present[i] <- FALSE
      asos_data$albedo[i]       <- alpha_soil
    }
  } else {
    asos_data$albedo[i]      <- alpha_soil
    asos_data$snow_present[i] <- FALSE
  }
}

ggplot(asos_data, aes(x = datetime, y = albedo)) +
  geom_line(color = "#2166ac", linewidth = 0.5) +
  geom_point(data = filter(asos_data, snowfall),
             aes(x = datetime, y = albedo),
             color = "red", size = 2, shape = 17) +
  geom_hline(yintercept = alpha_fresh, linetype = "dashed", color = "darkgreen", alpha = 0.5) +
  geom_hline(yintercept = alpha_decay_no_melt, linetype = "dashed", color = "orange", alpha = 0.5) +
  geom_hline(yintercept = alpha_decay_melt, linetype = "dashed", color = "red", alpha = 0.5) +
  geom_hline(yintercept = alpha_soil, linetype = "dashed", color = "brown", alpha = 0.5) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.1)) +
  labs(title = "Hourly Albedo Estimation — Telluride ASOS (WY2024)",
       subtitle = "Red triangles = snowfall events",
       x = "Date", y = "Albedo",
       caption = paste("Fresh snow α = 0.84 | No-melt decay min α = 0.70",
                       "| Melt decay min α = 0.50 | Soil α = 0.20")) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5))

cat("Mean albedo:", round(mean(asos_data$albedo, na.rm = TRUE), 3), "\n")
cat("Number of snowfall events detected:", sum(asos_data$snowfall, na.rm = TRUE), "\n")
cat("% of time with snow cover:",
    round(100 * sum(asos_data$snow_present) / nrow(asos_data), 1), "%\n")
```

## 2. Albedo and Temperature Relationship

```{r}
plot_data <- asos_data %>% mutate(temp_c = (tmpf - 32) * 5/9)

ggplot(plot_data, aes(x = datetime)) +
  geom_line(aes(y = albedo, color = "Albedo"), linewidth = 0.7) +
  geom_line(aes(y = temp_c / 50 + 0.5, color = "Temperature"),
            linewidth = 0.5, alpha = 0.7) +
  scale_y_continuous(
    name = "Albedo",
    sec.axis = sec_axis(~ (. - 0.5) * 50, name = "Temperature (°C)")
  ) +
  scale_color_manual(values = c("Albedo" = "#2166ac", "Temperature" = "#d73027")) +
  labs(title = "Albedo and Temperature — Telluride ASOS (WY2024)",
       x = "Date", color = "") +
  theme_minimal() +
  theme(legend.position = "bottom",
        plot.title = element_text(hjust = 0.5, face = "bold"))
```

## 3. Fresh Snow Density — Hedstrom-Pomeroy Equation

The Hedstrom-Pomeroy (1998) equation relates fresh snow density to air temperature:
ρ = 67.92 + 51.25 × exp(T / 2.59)

```{r}
calc_fresh_snow_density <- function(temp_c) {
  density <- 67.92 + 51.25 * exp(temp_c / 2.59)
  pmax(50, pmin(200, density))
}

snow_events <- asos_data %>%
  filter(!is.na(temp_c) & temp_c <= 0 & !is.na(p01m) & p01m > 0) %>%
  mutate(
    fresh_snow_density     = calc_fresh_snow_density(temp_c),
    hourly_swe_mm          = p01m,
    hourly_snow_depth_cm   = (hourly_swe_mm * (1000 / fresh_snow_density)) / 10,
    cumulative_snowfall_cm = cumsum(hourly_snow_depth_cm)
  )

cat("=== FRESH SNOW DENSITY SUMMARY ===\n")
cat(sprintf("Snow events: %d\n", nrow(snow_events)))
cat(sprintf("Mean fresh snow density: %.1f kg/m³\n", mean(snow_events$fresh_snow_density)))
cat(sprintf("Total cumulative snowfall: %.1f cm (%.1f in)\n",
            max(snow_events$cumulative_snowfall_cm),
            max(snow_events$cumulative_snowfall_cm) / 2.54))

ggplot(snow_events, aes(x = datetime, y = fresh_snow_density)) +
  geom_point(aes(color = temp_c), alpha = 0.6, size = 2) +
  geom_smooth(method = "loess", color = "black", se = TRUE, alpha = 0.2) +
  scale_color_gradient2(name = "Air Temp (°C)",
                        low = "#2166ac", mid = "#f7f7f7", high = "#b2182b",
                        midpoint = -8) +
  labs(title = "Hourly Fresh Snow Density During Snow Events (WY2024)",
       subtitle = "Telluride ASOS — Hedstrom-Pomeroy equation",
       x = "Date", y = "Fresh Snow Density (kg/m³)") +
  theme_minimal() +
  theme(legend.position = "bottom",
        plot.title = element_text(hjust = 0.5, face = "bold"))
```

## 4. Density–Temperature Relationship and Model Comparison

```{r}
SNOTEL <- read.table("data/SNOWTEL.txt", header = TRUE, sep = "\t")
colnames(SNOTEL) <- trimws(colnames(SNOTEL))
SNOTEL$Date     <- as.Date(SNOTEL$Date, format = "%m/%d/%Y")
SNOTEL           <- SNOTEL[order(SNOTEL$Date), ]

SNOTEL$delta_depth <- c(NA, diff(SNOTEL$SNWD_in))
SNOTEL$delta_swe   <- c(NA, diff(SNOTEL$WTEQ_in))
SNOTEL$fresh_snow_event <- !is.na(SNOTEL$delta_depth) &
  !is.na(SNOTEL$delta_swe) &
  SNOTEL$delta_depth >= 0.5 &
  SNOTEL$delta_swe >= 0.05

SNOTEL$fresh_snow_density_obs <- ifelse(
  SNOTEL$fresh_snow_event & SNOTEL$delta_depth > 0,
  (SNOTEL$delta_swe / SNOTEL$delta_depth) * 1000, NA)

daily_temp <- asos_data %>%
  mutate(date = as.Date(datetime)) %>%
  group_by(date) %>%
  summarise(avg_temp_c = mean(temp_c, na.rm = TRUE), .groups = "drop")

fresh_snow <- SNOTEL %>%
  filter(fresh_snow_event & !is.na(fresh_snow_density_obs) &
           fresh_snow_density_obs >= 50 & fresh_snow_density_obs <= 250) %>%
  select(Date, delta_depth, delta_swe, fresh_snow_density_obs) %>%
  left_join(daily_temp, by = c("Date" = "date")) %>%
  filter(!is.na(avg_temp_c))

if (nrow(fresh_snow) > 1) {
  model_linear <- lm(fresh_snow_density_obs ~ avg_temp_c, data = fresh_snow)
  temp_seq     <- seq(min(fresh_snow$avg_temp_c) - 2,
                      max(fresh_snow$avg_temp_c) + 2, length.out = 100)
  pred_df <- data.frame(
    temp_c          = temp_seq,
    SNOTEL_Linear   = predict(model_linear, data.frame(avg_temp_c = temp_seq)),
    Hedstrom_Pomeroy = calc_fresh_snow_density(temp_seq)
  )
  pred_long <- tidyr::pivot_longer(pred_df, -temp_c,
                                   names_to = "Model", values_to = "Density")

  ggplot() +
    geom_point(data = fresh_snow,
               aes(x = avg_temp_c, y = fresh_snow_density_obs),
               size = 3, alpha = 0.6) +
    geom_line(data = pred_long,
              aes(x = temp_c, y = Density, color = Model, linetype = Model),
              linewidth = 1.2) +
    scale_color_manual(values = c("SNOTEL_Linear" = "#4575b4",
                                  "Hedstrom_Pomeroy" = "#d73027")) +
    labs(title = "Fresh Snow Density vs. Air Temperature Calibration",
         subtitle = paste("SNOTEL Site 713 & Telluride ASOS —",
                          nrow(fresh_snow), "fresh snow events"),
         x = "Average Air Temperature (°C)", y = "Fresh Snow Density (kg/m³)",
         caption = paste("Linear R² =", round(summary(model_linear)$r.squared, 3))) +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"),
          legend.position = "bottom")
}
```
