---
title: "Blowing Snow: Snow Metamorphism, Densification, and Pomeroy Transport"
author: "Karen Dworsky"
date: "2025-11-10"
format: html
execute:
  cache: true
---

## Setup

```{r}
library(dplyr)
library(ggplot2)
library(lubridate)
library(tidyr)

asos_data <- read.csv("data/ASOS2.txt", header = TRUE, stringsAsFactors = FALSE)

asos_data <- asos_data %>%
  mutate(
    datetime   = as.POSIXct(valid, format = "%Y-%m-%d %H:%M", tz = "UTC"),
    p01i       = as.numeric(ifelse(p01i %in% c("M", ""), NA, p01i)),
    tmpf       = as.numeric(ifelse(tmpf %in% c("M", ""), NA, tmpf)),
    sknt       = as.numeric(ifelse(sknt %in% c("M", ""), NA, sknt)),
    temp_C     = (tmpf - 32) * 5/9,
    snowfall_m = ifelse(!is.na(temp_C) & !is.na(p01i) & temp_C <= 0 & p01i > 0,
                        p01i * 0.0254, 0)
  ) %>%
  arrange(datetime)
```

## 1. Snow Metamorphism and Densification Model

The densification model tracks old and new snow layers separately:
- Old snow density approaches ρ_max via exponential decay
- New snow density uses the temperature-dependent equation: ρ = 50 + 1.7 × (T + 15)^(3/2)

```{r}
rho_max     <- 350   # kg/m³ — maximum density (from SNOTEL observations)
k           <- 0.01  # densification rate (per hour)
density_old <- 250   # kg/m³ — initial old snow density
depth_old   <- 0
depth_new   <- 0
density_new <- 0

depth_total   <- numeric(nrow(asos_data))
density_total <- numeric(nrow(asos_data))

new_snow_density <- function(temp_C) {
  pmax(50, pmin(400, 50 + 1.7 * (temp_C + 15)^(3/2)))
}

for (i in seq_len(nrow(asos_data))) {
  temp_C     <- asos_data$temp_C[i]
  snowfall_m <- asos_data$snowfall_m[i]

  density_old <- rho_max - (rho_max - density_old) * exp(-k)
  depth_old   <- depth_old * density_old / (density_old + k * depth_old)

  if (!is.na(temp_C) && !is.na(snowfall_m) && snowfall_m > 0) {
    density_new <- new_snow_density(temp_C)
    depth_new   <- depth_new + snowfall_m * 100  # convert m → cm
  }

  total_depth <- depth_old + depth_new
  depth_total[i] <- total_depth
  if (total_depth > 0) {
    density_total[i] <- (depth_old * density_old + depth_new * density_new) / total_depth
  } else {
    density_total[i] <- density_old
  }

  if (!is.na(snowfall_m) && snowfall_m > 0) {
    depth_old   <- depth_total[i]
    density_old <- density_total[i]
    depth_new   <- 0
    density_new <- 0
  }
}

asos_data$modeled_depth_cm   <- depth_total
asos_data$modeled_density    <- density_total

asos_data <- asos_data %>%
  mutate(
    snowfall_cm        = snowfall_m * 100,
    cumulative_snowfall_cm = cumsum(snowfall_cm)
  )
```

## 2. Modeled Snow Depth vs. Cumulative Snowfall

```{r}
depth_plot <- ggplot(asos_data, aes(x = datetime)) +
  geom_line(aes(y = cumulative_snowfall_cm, color = "Cumulative Snowfall"), linewidth = 1) +
  geom_line(aes(y = modeled_depth_cm, color = "Modeled Depth (with metamorphism)"),
            linewidth = 1) +
  scale_color_manual(
    values = c("Cumulative Snowfall" = "#2166ac",
               "Modeled Depth (with metamorphism)" = "#1a9641")) +
  labs(title = "Cumulative Snowfall vs. Modeled Snow Depth (WY2024)",
       subtitle = "Telluride ASOS — Snow densification model",
       x = "Date", y = "Snow Depth (cm)", color = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom",
        plot.title = element_text(hjust = 0.5, face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1))

print(depth_plot)
```

## 3. Hourly Bulk Snow Density

```{r}
density_plot <- ggplot(asos_data, aes(x = datetime, y = modeled_density)) +
  geom_line(color = "#4575b4") +
  labs(title = "Hourly Bulk Snow Density (WY2024)",
       subtitle = "Telluride ASOS — Exponential densification model",
       x = "Date", y = expression("Density (kg/m"^3*")")) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1))

print(density_plot)
```

## 4. Hedstrom-Pomeroy Fresh Snow Density

```{r}
hedstrom_pomeroy <- function(temp_C) {
  pmax(50, pmin(200, 67.92 + 51.25 * exp(temp_C / 2.59)))
}

asos_data <- asos_data %>%
  mutate(fresh_snow_density_hp = ifelse(!is.na(temp_C) & temp_C <= 0 & !is.na(p01i) & p01i > 0,
                                        hedstrom_pomeroy(temp_C), NA))

ggplot(asos_data, aes(x = datetime, y = fresh_snow_density_hp)) +
  geom_line(color = "#4575b4") +
  labs(title = "Hourly Fresh Snow Density — Hedstrom-Pomeroy (WY2024)",
       subtitle = "Telluride ASOS",
       x = "Date", y = "Fresh Snow Density (kg/m³)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

asos_monthly <- asos_data %>%
  mutate(month = floor_date(datetime, "month")) %>%
  group_by(month) %>%
  summarise(avg_density = mean(fresh_snow_density_hp, na.rm = TRUE), .groups = "drop")

ggplot(asos_monthly, aes(x = month, y = avg_density)) +
  geom_col(fill = "#4575b4", alpha = 0.8) +
  scale_x_datetime(date_labels = "%b %Y") +
  labs(title = "Monthly Average Fresh Snow Density — Hedstrom-Pomeroy (WY2024)",
       x = "Month", y = "Average Fresh Snow Density (kg/m³)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5, face = "bold"))
```

## 5. Observed vs. Modeled Bulk Density (SNOTEL) — Sturm et al. 2010

```{r}
snotel_data <- read.table("data/SNOWTEL.txt", header = TRUE, sep = "\t")
colnames(snotel_data) <- trimws(colnames(snotel_data))
snotel_data$Date  <- as.Date(snotel_data$Date, format = "%m/%d/%Y")
snotel_data       <- snotel_data[order(snotel_data$Date), ]

snotel_data <- snotel_data %>%
  mutate(
    depth_m     = SNWD_in * 0.0254,
    swe_m       = WTEQ_in * 0.0254,
    obs_density = ifelse(depth_m > 0, swe_m / depth_m, NA),
    doy         = row_number()
  )

sturm_density <- function(depth_cm, doy) {
  rho_max <- 0.5975
  k1      <- 0.0012
  k2      <- 0.0038
  rho_0   <- 0.024
  rho_max * (1 - exp(-k1 * depth_cm - k2 * doy)) + rho_0
}

snotel_data <- snotel_data %>%
  mutate(sturm_density_gcc = sturm_density(depth_m * 100, doy))

ggplot(snotel_data, aes(x = Date)) +
  geom_point(aes(y = obs_density), color = "#2166ac", size = 2, alpha = 0.7) +
  geom_line(aes(y = sturm_density_gcc), color = "#d73027", linewidth = 1) +
  labs(title = "Observed vs. Modeled Bulk Snow Density (WY2024)",
       subtitle = "Red Mountain Pass SNOTEL Site 713 | Sturm et al. 2010 (alpine)",
       caption = "Observed (blue) | Modeled (red)",
       x = "Date", y = "Bulk Density (g/cm³)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```
