---
title: "Radiation & Energy Balance: Longwave, Shortwave, and Net Energy"
author: "Karen Dworsky"
date: "2025-11-14"
format: html
execute:
  cache: true
---

## Setup

All radiation components were computed from the Telluride ASOS (TEX) station WY2024 data.
Longwave radiation used an all-sky emissivity approach; shortwave used a solar geometry model
accounting for cloud cover fraction.

```{r}
library(dplyr)
library(ggplot2)
library(lubridate)
library(tidyr)

data <- read.csv("data/data_radiation1.csv", stringsAsFactors = FALSE)

data <- data %>%
  mutate(
    datetime = as.POSIXct(datetime, format = "%m/%d/%Y", tz = "UTC"),
    lw_net   = as.numeric(lw_net),
    HK_in    = as.numeric(HK_in),
    HK_star  = as.numeric(HK_star),
    Qcum     = as.numeric(Qcum)
  ) %>%
  filter(!is.na(datetime))
```

## 1. Net Longwave Radiation

All-sky emissivity (Brutsaert 1975 modified):
ε_all = 0.83 − 0.18 × exp(−0.067 × √ea)

Net longwave = ε_all × σT⁴ − σT⁴ = (ε_all − 1) × σT⁴

```{r}
ggplot(data, aes(x = datetime, y = lw_net)) +
  geom_line(color = "#d73027", linewidth = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  labs(title = "Net Longwave Radiation — Telluride ASOS (WY2024)",
       x = "Date", y = "Net Longwave Radiation (W/m²)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

## 2. Incoming and Net Shortwave Radiation

Incoming shortwave computed from solar geometry (Spencer 1971, Spencer 1971 equation of time),
attenuated by cloud cover fraction using the Kasten-Young cloudiness transmissivity:
HK_in = I_sc × E₀ × cos(Z) × (0.355 + 0.68 × (1 − C_cloud))

```{r}
sw_plot <- data %>%
  filter(!is.na(HK_in) & !is.na(HK_star)) %>%
  select(datetime, HK_in, HK_star) %>%
  pivot_longer(cols = c(HK_in, HK_star),
               names_to = "variable", values_to = "W_m2")

ggplot(sw_plot, aes(x = datetime, y = W_m2, color = variable)) +
  geom_line(linewidth = 0.6, alpha = 0.8) +
  scale_color_manual(name = NULL,
                     values = c("HK_in" = "#fdae61", "HK_star" = "#2166ac"),
                     labels = c("HK_in" = "Incoming Shortwave",
                                "HK_star" = "Net Shortwave")) +
  labs(title = "Shortwave Radiation — Telluride ASOS (WY2024)",
       x = "Date", y = "Shortwave Radiation (W/m²)") +
  theme_minimal() +
  theme(legend.position = "bottom",
        plot.title = element_text(hjust = 0.5, face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

## 3. Cumulative Net Energy Balance

The cumulative energy balance combines net longwave, net shortwave, and energy added
by precipitation (rain-on-snow heat flux).

```{r}
ggplot(data, aes(x = datetime, y = Qcum)) +
  geom_line(color = "#4575b4", linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  labs(title = "Cumulative Net Energy Balance — Telluride ASOS (WY2024)",
       subtitle = "Longwave + shortwave + precipitation energy",
       x = "Date", y = "Cumulative Net Energy (J/m²)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

## 4. Monthly Average Radiation Components

```{r}
monthly_rad <- data %>%
  mutate(month = floor_date(datetime, "month")) %>%
  group_by(month) %>%
  summarise(
    mean_lw_net  = mean(lw_net,  na.rm = TRUE),
    mean_HK_star = mean(HK_star, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_longer(cols = c(mean_lw_net, mean_HK_star),
               names_to = "component", values_to = "W_m2")

ggplot(monthly_rad, aes(x = month, y = W_m2, fill = component)) +
  geom_col(position = "dodge", alpha = 0.85) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_fill_manual(name = "Component",
                    values = c("mean_lw_net" = "#d73027", "mean_HK_star" = "#2166ac"),
                    labels = c("mean_lw_net" = "Net Longwave",
                               "mean_HK_star" = "Net Shortwave")) +
  scale_x_datetime(date_labels = "%b %Y") +
  labs(title = "Monthly Average Radiation Components — Telluride ASOS (WY2024)",
       x = "Month", y = "Average Radiation (W/m²)") +
  theme_minimal() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5, face = "bold"))
```

## 5. Summary Statistics

```{r}
cat("=== RADIATION SUMMARY (WY2024) ===\n")
cat(sprintf("Mean net longwave:    %.1f W/m²\n", mean(data$lw_net,  na.rm = TRUE)))
cat(sprintf("Mean incoming SW:     %.1f W/m²\n", mean(data$HK_in,   na.rm = TRUE)))
cat(sprintf("Mean net SW:          %.1f W/m²\n", mean(data$HK_star, na.rm = TRUE)))
cat(sprintf("Final cumulative Q:   %.0f J/m²\n",  tail(data$Qcum[!is.na(data$Qcum)], 1)))
```
