---
title: "Snowmelt & Runoff: Melt Factor Calibration and Baseflow Separation"
author: "Karen Dworsky"
date: "2025-11-19"
format: html
execute:
  cache: true
---

## Setup

```{r}
library(dplyr)
library(ggplot2)
library(lubridate)
library(tidyr)
```

## 1. Melt Factor Calibration

A temperature-index melt equation was calibrated using SNOTEL SWE change data:
M = MF × (T − T₀), where T > T₀

Non-linear least squares estimates MF (melt factor) and T₀ (base temperature).

```{r}
swe_data <- read.csv("data/NWIS_snowmelt.csv", stringsAsFactors = FALSE)

swe_data <- swe_data %>%
  mutate(datetime = mdy_hm(datetime))

cat("=== DATA OVERVIEW ===\n")
cat(sprintf("Records: %d\n", nrow(swe_data)))
cat(sprintf("Temperature range: %.1f to %.1f °C\n",
            min(swe_data$temp_C, na.rm = TRUE),
            max(swe_data$temp_C, na.rm = TRUE)))
cat(sprintf("SWE change range: %.2f to %.2f mm\n",
            min(swe_data$change_SWE, na.rm = TRUE),
            max(swe_data$change_SWE, na.rm = TRUE)))

ggplot(swe_data, aes(x = temp_C, y = change_SWE)) +
  geom_point(alpha = 0.5, color = "#4575b4") +
  labs(title = "SWE Change vs. Temperature",
       subtitle = "SNOTEL Site 713 — Red Mountain Pass",
       x = "Air Temperature (°C)", y = "SWE Change (mm)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

## 2. Fit Temperature-Index Melt Equation

```{r}
swe_filtered <- swe_data %>%
  filter(change_SWE != 0 & !is.na(temp_C) & !is.na(change_SWE) & temp_C <= 13)

melt_model <- nls(change_SWE ~ MF * (temp_C - T0),
                  data    = swe_filtered,
                  start   = list(MF = 0.05, T0 = 5),
                  control = nls.control(maxiter = 500))

mf_estimate <- coef(melt_model)[["MF"]]
t0_estimate <- coef(melt_model)[["T0"]]

cat(sprintf("Melt factor (MF): %.4f mm/°C\n", mf_estimate))
cat(sprintf("Base temperature (T0): %.2f °C\n", t0_estimate))
cat(sprintf("Melt equation: M = %.4f × (T − %.2f)\n", mf_estimate, t0_estimate))

temp_seq <- seq(t0_estimate, 13, length.out = 100)
fit_line  <- data.frame(temp_C = temp_seq,
                         change_SWE = mf_estimate * (temp_seq - t0_estimate))

ggplot(swe_filtered, aes(x = temp_C, y = change_SWE)) +
  geom_point(alpha = 0.5, color = "#4575b4") +
  geom_line(data = fit_line, color = "#d73027", linewidth = 1.2) +
  labs(title = "Melt Factor Calibration — Temperature-Index Model",
       subtitle = sprintf("M = %.4f × (T − %.2f°C)", mf_estimate, t0_estimate),
       x = "Temperature (°C)", y = "SWE Change (mm)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

## 3. Baseflow Separation — Lynne-Hollick Digital Filter

The Lynne-Hollick recursive digital filter separates streamflow into baseflow and
quickflow components. Three passes improve baseflow isolation.

```{r}
discharge_data <- read.csv("data/NWIS_discharge.csv", stringsAsFactors = FALSE)

discharge_data <- discharge_data %>%
  mutate(date     = mdy(time),
         flow_cfs = as.numeric(value)) %>%
  filter(!is.na(flow_cfs)) %>%
  arrange(date)

lynne_hollick_filter <- function(Q, a = 0.925) {
  bf    <- numeric(length(Q))
  bf[1] <- Q[1]
  for (i in 2:length(Q)) {
    bf[i] <- ((1 - a) / (1 + a)) * Q[i] + (a / (1 + a)) * bf[i - 1]
    bf[i] <- min(bf[i], Q[i])
  }
  bf
}

baseflow <- discharge_data$flow_cfs
for (i in 1:3) {
  baseflow <- lynne_hollick_filter(baseflow)
}

discharge_data$baseflow_cfs   <- baseflow
discharge_data$quickflow_cfs  <- discharge_data$flow_cfs - baseflow

total_volume    <- sum(discharge_data$flow_cfs)    * 86400 / 43559.9
baseflow_volume <- sum(discharge_data$baseflow_cfs) * 86400 / 43559.9

cat("=== BASEFLOW SEPARATION RESULTS ===\n")
cat(sprintf("Total flow volume:    %.0f acre-feet\n", total_volume))
cat(sprintf("Baseflow volume:      %.0f acre-feet\n", baseflow_volume))
cat(sprintf("Baseflow fraction:    %.1f%%\n", baseflow_volume / total_volume * 100))
```

## 4. Hydrograph Separation Plot

```{r}
hydro_long <- discharge_data %>%
  select(date, Total = flow_cfs, Baseflow = baseflow_cfs) %>%
  pivot_longer(cols = c(Total, Baseflow), names_to = "component", values_to = "flow_cfs")

ggplot(discharge_data, aes(x = date)) +
  geom_ribbon(aes(ymin = baseflow_cfs, ymax = flow_cfs),
              fill = "#74add1", alpha = 0.5) +
  geom_line(aes(y = flow_cfs,    color = "Total Flow"),  linewidth = 1) +
  geom_line(aes(y = baseflow_cfs, color = "Baseflow"),    linewidth = 1) +
  scale_color_manual(name = NULL,
                     values = c("Total Flow" = "black", "Baseflow" = "#2166ac")) +
  labs(title = "Hydrograph Separation — Lynne-Hollick Filter (WY2024)",
       subtitle = "Shaded area = quickflow (snowmelt runoff component)",
       x = "Date", y = "Flow (cfs)") +
  theme_minimal() +
  theme(legend.position = "bottom",
        plot.title = element_text(hjust = 0.5, face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

## 5. Hourly Melt Estimation

Applying the calibrated melt factor to hourly ASOS temperature data:

```{r}
asos_melt <- read.csv("data/asos_data.txt", stringsAsFactors = FALSE)

asos_melt <- asos_melt %>%
  mutate(
    valid  = ymd_hm(valid),
    tmpc   = as.numeric(tmpc),
    hourly_melt = ifelse(!is.na(tmpc) & tmpc > t0_estimate,
                         mf_estimate * (tmpc - t0_estimate), 0)
  ) %>%
  filter(!is.na(valid))

cat(sprintf("Total estimated melt: %.1f mm\n", sum(asos_melt$hourly_melt, na.rm = TRUE)))

asos_daily_melt <- asos_melt %>%
  mutate(date = as.Date(valid)) %>%
  group_by(date) %>%
  summarise(daily_melt = sum(hourly_melt, na.rm = TRUE), .groups = "drop")

ggplot(asos_daily_melt, aes(x = date, y = daily_melt)) +
  geom_col(fill = "#4575b4", alpha = 0.8) +
  labs(title = "Daily Estimated Snowmelt — Temperature-Index Model (WY2024)",
       subtitle = sprintf("MF = %.4f mm/°C, T₀ = %.1f°C", mf_estimate, t0_estimate),
       x = "Date", y = "Daily Melt (mm)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1))
```
