---
title: "Snow Interception: Forest Canopy Interception by Tree Species"
author: "Karen Dworsky"
date: "2025-11-10"
format: html
execute:
  cache: true
---

## Setup

Forest canopy intercepts a fraction of snowfall depending on leaf area index (LAI) and
temperature. This analysis applies the LAI-based interception model to three tree species
common in the Telluride area using WY2024 ASOS data.

```{r}
library(dplyr)
library(ggplot2)
library(lubridate)
library(tidyr)
library(zoo)

ASOS <- read.table("data/ASOS2.txt", sep = ",", header = TRUE)

ASOS <- ASOS %>%
  mutate(
    valid  = as.POSIXct(valid, format = "%Y-%m-%d %H:%M", tz = "UTC"),
    p01i   = as.numeric(ifelse(p01i == "M" | p01i == "" | is.na(p01i), NA, p01i)),
    tmpf   = as.numeric(ifelse(tmpf == "M" | tmpf == "" | is.na(tmpf), NA, tmpf))
  ) %>%
  arrange(valid)
```

## 1. LAI Values and Interception Model

LAI values for three tree species common in the San Juan Mountains (summer/winter):

```{r}
LAI_quaking_aspen_summer  <- 3.0
LAI_quaking_aspen_winter  <- 0.0
LAI_douglas_fir_summer    <- 6.0
LAI_douglas_fir_winter    <- 5.0
LAI_western_larch_summer  <- 2.0
LAI_western_larch_winter  <- 0.0

get_season <- function(date) {
  ifelse(month(date) >= 6 & month(date) <= 8, "summer", "winter")
}

calculate_interception <- function(precipitation, temperature, lai_summer, lai_winter, valid_date) {
  lai <- ifelse(get_season(valid_date) == "summer", lai_summer, lai_winter)
  ifelse(temperature < 32,
         0.6 * lai * precipitation,
         0.2 * lai * precipitation)
}

ASOS <- ASOS %>%
  mutate(
    p01i_clean = ifelse(is.na(p01i), 0, p01i),
    tmpf_clean = ifelse(is.na(tmpf), 32, tmpf),
    interception_quaking_aspen = mapply(calculate_interception,
      p01i_clean, tmpf_clean, LAI_quaking_aspen_summer, LAI_quaking_aspen_winter, valid),
    interception_douglas_fir = mapply(calculate_interception,
      p01i_clean, tmpf_clean, LAI_douglas_fir_summer, LAI_douglas_fir_winter, valid),
    interception_western_larch = mapply(calculate_interception,
      p01i_clean, tmpf_clean, LAI_western_larch_summer, LAI_western_larch_winter, valid),
    is_storm = p01i_clean > 0
  )

ASOS <- ASOS %>%
  mutate(
    storm_end = ifelse(is_storm, as.numeric(valid), NA_real_),
    storm_end = zoo::na.locf(storm_end, na.rm = FALSE),
    storm_end = as.POSIXct(storm_end, origin = "1970-01-01", tz = "UTC"),
    hours_since_storm = as.numeric(difftime(valid, storm_end, units = "hours")),
    interception_quaking_aspen = ifelse(is.na(storm_end) | hours_since_storm > 6,
                                        0, interception_quaking_aspen),
    interception_douglas_fir   = ifelse(is.na(storm_end) | hours_since_storm > 6,
                                        0, interception_douglas_fir),
    interception_western_larch = ifelse(is.na(storm_end) | hours_since_storm > 6,
                                        0, interception_western_larch)
  )
```

## 2. Monthly Average Interception

```{r}
monthly_avg <- ASOS %>%
  mutate(month = floor_date(valid, "month")) %>%
  group_by(month) %>%
  summarise(
    quaking_aspen  = mean(interception_quaking_aspen, na.rm = TRUE),
    douglas_fir    = mean(interception_douglas_fir, na.rm = TRUE),
    western_larch  = mean(interception_western_larch, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_longer(cols = c(quaking_aspen, douglas_fir, western_larch),
               names_to = "species", values_to = "avg_interception")

ggplot(monthly_avg, aes(x = month, y = avg_interception, color = species, group = species)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_color_manual(name = "Tree Species",
                     values = c("quaking_aspen" = "#4575b4",
                                "douglas_fir"   = "#1a9641",
                                "western_larch" = "#d73027"),
                     labels = c("quaking_aspen" = "Quaking Aspen",
                                "douglas_fir"   = "Douglas Fir",
                                "western_larch" = "Western Larch")) +
  scale_x_datetime(date_labels = "%b %Y") +
  labs(title = "Monthly Average Canopy Interception by Tree Species (WY2024)",
       subtitle = "Telluride ASOS — LAI-based interception model",
       x = "Month", y = "Average Interception (inches)") +
  theme_minimal() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5, face = "bold"))
```

## 3. Interception During Storm Events

```{r}
monthly_storm <- ASOS %>%
  filter(is_storm) %>%
  mutate(month = floor_date(valid, "month")) %>%
  group_by(month) %>%
  summarise(
    quaking_aspen  = mean(interception_quaking_aspen, na.rm = TRUE),
    douglas_fir    = mean(interception_douglas_fir, na.rm = TRUE),
    western_larch  = mean(interception_western_larch, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_longer(cols = c(quaking_aspen, douglas_fir, western_larch),
               names_to = "species", values_to = "storm_interception")

ggplot(monthly_storm, aes(x = month, y = storm_interception, color = species, group = species)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_color_manual(name = "Tree Species",
                     values = c("quaking_aspen" = "#4575b4",
                                "douglas_fir"   = "#1a9641",
                                "western_larch" = "#d73027"),
                     labels = c("quaking_aspen" = "Quaking Aspen",
                                "douglas_fir"   = "Douglas Fir",
                                "western_larch" = "Western Larch")) +
  scale_x_datetime(date_labels = "%b %Y") +
  labs(title = "Average Canopy Interception During Storm Events (WY2024)",
       subtitle = "Telluride ASOS — Storm hours only",
       x = "Month", y = "Average Interception (inches)") +
  theme_minimal() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5, face = "bold"))
```

## 4. Summary Statistics

```{r}
total_precip <- sum(ASOS$p01i_clean, na.rm = TRUE)

species_totals <- ASOS %>%
  summarise(
    total_aspen  = sum(interception_quaking_aspen, na.rm = TRUE),
    total_fir    = sum(interception_douglas_fir,   na.rm = TRUE),
    total_larch  = sum(interception_western_larch,  na.rm = TRUE)
  )

cat("=== ANNUAL INTERCEPTION SUMMARY (WY2024) ===\n")
cat(sprintf("Total precipitation: %.2f inches\n", total_precip))
cat(sprintf("Quaking Aspen: %.3f in (%.1f%%)\n",
            species_totals$total_aspen,
            species_totals$total_aspen / total_precip * 100))
cat(sprintf("Douglas Fir:   %.3f in (%.1f%%)\n",
            species_totals$total_fir,
            species_totals$total_fir / total_precip * 100))
cat(sprintf("Western Larch: %.3f in (%.1f%%)\n",
            species_totals$total_larch,
            species_totals$total_larch / total_precip * 100))
```
